Does being on the podium affect Formula 1 drivers' performance in the next race?

### 2025-01-30 inst/podium.Rmd
magrittr package needed
more detail

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
options("digits" = 5)
```

@@

To cite the `hyper2` package in publications, please use @hankin2017_rmd.
This short document defines the `podium()` function which tests the
hypothesis that a Formula 1 driver's being on the podium at one venue
affects his performance on the next.  Data from the 2007 season are
used.  It uses `hyper3` objects from the `hyper2` package.  A detailed
use-case for the top six drivers (Alonso, Hamilton, Heidfeld, Kubica,
Massa, and Raikkonen), and only the first five venues (Austria,
Malaysia, Bahrain, Spain, and Monaco) is given.

@@

Define a subset of it:

@@

```{r,usepodiu,cache=TRUE}
Ha <- podium(1.4,b)
Hb <- podium(1.5,b)
ma <- maxp(Ha,give=TRUE,n=1)
mb <- maxp(Hb,give=TRUE,n=1)
```

```{r}
Ha
Hb
ma
mb
```

@@

```{r,podlikevec7,cache=TRUE}
date()
lam <- seq(from=0.2,to=1.2,len=7)
like <- lapply(lam,podlike)
date()
```

>>>>

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

@@

<font size="1"> (almost all the time is spent on chunk `podlikevec7`,
which takes about one minute per evaluation) </font>

To cite the `hyper2` package in publications, please use
@hankin2017_rmd.  This short document defines the `podium()` function
which tests the hypothesis that a Formula 1 driver's being on the
podium at one venue affects his performance on the next.  Data from
the 2007 season are used.  It uses `hyper3` objects from the `hyper2`
package.  A detailed use-case for the top six drivers (Alonso,
Hamilton, Heidfeld, Kubica, Massa, and Raikkonen^[Umlauts pose special
problems for Rmd files]), and only the first five venues (Austria,
Malaysia, Bahrain, Spain, and Monaco) is given.

@@

The first race has standard Plackett-Luce likelihood:

$$a\succ b\succ c\succ d\succ e\succ f$$

has a PL likelihood of

$$
\frac{a}{a+b+c+d+e+f}\cdot
\frac{b}{  b+c+d+e+f}\cdot
\frac{c}{    c+d+e+f}\cdot
\frac{d}{      d+e+f}\cdot
\frac{e}{        e+f}
$$

However, if the result were

$$\overline{a}\succ b\succ\overline{c}\succ\overline{d}\succ e\succ f$$

(where an overbar signifies that the driver was on the podium after
the previous race), the likelihood function would be

$$
\frac{\lambda a}{\lambda a+b+\lambda c+\lambda d+e+f}\cdot
\frac{        b}{          b+\lambda c+\lambda d+e+f}\cdot
\frac{\lambda c}{            \lambda c+\lambda d+e+f}\cdot
\frac{\lambda d}{                      \lambda d+e+f}\cdot
\frac{        e}{                                e+f}\cdot
$$

and the null of no podium effect would be $H_0\colon\lambda=0$.  Next,
we will take dataset `formula1_2007` and define a subset of it which
we will call `b`:

@@

Taking Monaco (`MON`) as an example, the preceding race was in spain
(`ESP`) at which Massa, Hamilton, and Alonso finished on the podium.
The order statistic would be

$$
\text{Raikkonen}\succ
\overline{\text{Hamilton}}\succ
\overline{\text{Massa}}\succ
\overline{\text{Alonso}}\succ
\text{Kubica}\succ
\text{Heidfeld}
$$

where again an overbar signifies that the driver was on the podium in
the previous race.  The likelihood function for Monaco would thus be

$$
\begin{split}
\operatorname{\mathcal{L}}\left(\lambda;
p_\text{Raikkonen},\ldots,p_\text{Heidfeld}
\right)
& =
\frac{p_\text{Raikkonen}}{p_\text{Raikkonen}+\lambda p_\text{Hamilton}+\lambda p_\text{Massa}+\lambda p_\text{Alonso}+p_\text{Kubica}+p_\text{Heidfeld}}\\
&{\qquad} \cdot\frac{\lambda p_\text{Hamilton}}{\lambda  p_\text{Hamilton}+\lambda p_\text{Massa}+\lambda p_\text{Alonso}+p_\text{Kubica}+p_\text{Heidfeld}}\\
&{\qquad}\cdot\frac{\lambda p_\text{Massa}}{\lambda p_\text{Massa}+\lambda p_\text{Alonso}+p_\text{Kubica}+p_\text{Heidfeld}}\\
&{\qquad}\cdot\frac{\lambda p_\text{Alonso}}{\lambda p_\text{Alonso}+p_\text{Kubika}+p_\text{Heidfeld}}\cdot
\frac{p_\text{Kubica}}{p_\text{Kubica}+p_\text{Heidfeld}}
\end{split}
$$

```{r,usepodiu,cache=TRUE}
Ha <- podium(1.4, b)
Hb <- podium(2.1, b)
ma <- maxp(Ha, give=0, n=1)
mb <- maxp(Hb, give=0, n=1)
```

```{r}
Ha
summary(Hb)
ma
mb
```

Above we see that passing $\lambda$ values of 1.4 and 2.1 result in
different likelihoods.  The difference of about 0.4 units of support
is not significant, failing to reach the two-units-of-support
criterion of Edwards.

@@

```{r,podlikevec7,cache=TRUE}
date()
lam <- seq(from=0.2,to=1.2,len=15)
like <- lapply(lam,podlike)
date()
```
